---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "70%"
)
```

# rsgl

<!-- badges: start -->
[![R-CMD-check](https://github.com/sgl-projects/rsgl/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sgl-projects/rsgl/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

rsgl implements the [SGL graphics language](https://arxiv.org/pdf/2505.14690) for use within R.
SGL is a graphics language that is designed to look and feel like SQL,
and is based on the grammar of graphics.

## Installation

rsgl is not yet available on CRAN but can be installed from GitHub with:
```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("sgl-projects/rsgl")
```

## Usage
`dbGetPlot` is the primary interface to rsgl.
It takes a [DBI](https://dbi.r-dbi.org) database connection
and a SGL statement and returns a [ggplot2](https://ggplot2.tidyverse.org) plot.

The following example demonstrates creating a DBI connection to an in-memory
[DuckDB](https://duckdb.org) database, loading it with data,
and then generating a scatterplot from a SGL statement.

```{r usage}
library(duckdb)
library(rsgl)

con <- dbConnect(duckdb())
dbWriteTable(con, "cars", mtcars)

dbGetPlot(con, "
	visualize
		hp as x,
		mpg as y
	from cars
	using points
")
```