---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "70%"
)
```

# rsgl

<!-- badges: start -->
[![R-CMD-check](https://github.com/sgl-projects/rsgl/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sgl-projects/rsgl/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

rsgl implements the [SGL graphics language](https://arxiv.org/pdf/2505.14690) for use within R.
SGL is a graphics language that is designed to look and feel like SQL,
and is based on the grammar of graphics.

## Installation

To install the official [CRAN](https://cran.r-project.org) release:

```{r, eval=FALSE}
install.packages("rsgl")
```

If you instead want the development version, install from GitHub with:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("sgl-projects/rsgl")
```

## Usage
`dbGetPlot` is the primary interface to rsgl.
It takes a [DBI](https://dbi.r-dbi.org) database connection
and a SGL statement and returns the corresponding plot.

The following example demonstrates creating a DBI connection to an in-memory
[DuckDB](https://duckdb.org) database, loading it with data,
and then generating a scatterplot from a SGL statement.

```{r usage, warning = FALSE}
library(duckdb)
library(rsgl)

con <- dbConnect(duckdb())
dbWriteTable(con, "cars", cars)

dbGetPlot(con, "
	visualize
		horsepower as x,
		miles_per_gallon as y
	from cars
	using points
")
```

## Next steps

- [Get started](https://sgl-projects.github.io/rsgl/articles/rsgl.html) — a tutorial of rsgl and the SGL language.
- [Example gallery](https://sgl-projects.github.io/rsgl/articles/example-gallery.html) — a collection of plots generated with rsgl.