---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "70%"
)
```

# rsgl

<!-- badges: start -->
[![R-CMD-check](https://github.com/sgl-projects/rsgl/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sgl-projects/rsgl/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

rsgl implements [SGL (Structured Graphics Language)](https://arxiv.org/pdf/2505.14690), a declarative language for specifying statistical graphics that is designed to feel like SQL. You write a SGL statement, pass it to `dbGetPlot()` with a DuckDB connection, and get back a ggplot2 plot.

## Installation

You can install rsgl from GitHub with:

```r
# install.packages("remotes")
remotes::install_github("sgl-projects/rsgl")
```

## Examples

```{r readme-setup, include = FALSE}
library(duckdb)
library(rsgl)

con <- dbConnect(duckdb())
dbWriteTable(con, "cars", mtcars)
```

### Scatterplot

```{r scatterplot}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y,
    cyl as color
  from cars
  using points
")
```

### Histogram

```{r histogram}
dbGetPlot(con, "
  visualize
    bin(mpg) as x,
    count(*) as y
  from cars
  group by
    bin(mpg)
  using bars
")
```

### Scatterplot with regression line

```{r layered}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from cars
  using (
    points
    layer
    regression line
  )
  scale by
    log(x)
")
```

### Faceted box plot

```{r faceted}
dbGetPlot(con, "
  visualize
    cyl_cat as x,
    mpg as y
  from (
    select mpg, cast(cyl as varchar) as cyl_cat, am
    from cars
  )
  using boxes
  facet by
    am
")
```

```{r readme-cleanup, include = FALSE}
DBI::dbDisconnect(con, shutdown = TRUE)
```

## Learn more

- `vignette("rsgl")` — Get started with rsgl
- `vignette("sgl-language-guide")` — Full SGL syntax reference
