---
title: "Example Gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example Gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

A collection of SGL examples organized by plot type. Each example is a
self-contained SGL statement you can copy and adapt. For a full syntax
reference, see `vignette("sgl-language-guide")`.

```{r setup, include = FALSE}
library(rsgl)
library(duckdb)

con <- dbConnect(duckdb())
dbWriteTable(con, "cars", mtcars)
dbWriteTable(con, "trees", as.data.frame(Orange))
dbWriteTable(con, "diamonds", ggplot2::diamonds)
```

## Scatterplots

### Basic scatterplot

```{r scatter-basic}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from cars
  using points
")
```

### Colored by group

```{r scatter-color}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y,
    cyl as color
  from cars
  using points
")
```

### Sized by variable

```{r scatter-size}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y,
    wt as size
  from cars
  using points
")
```

### With regression line

```{r scatter-regression}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from cars
  using (
    points
    layer
    regression line
  )
")
```

### Jittered points

```{r scatter-jittered}
set.seed(42)
dbGetPlot(con, "
  visualize
    cyl_cat as x,
    mpg as y
  from (
    select mpg, cast(cyl as varchar) as cyl_cat
    from cars
  )
  using jittered points
")
```

### Filtered with SQL subquery

```{r scatter-subquery}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y,
    cyl as color
  from (
    select
      hp, mpg,
      cyl::varchar as cyl
    from cars
    where cyl = 4 or cyl = 6
  )
  using points
")
```

## Bar charts

### Counts per category

```{r bar-count}
dbGetPlot(con, "
  visualize
    cut as x,
    count(*) as y
  from diamonds
  group by
    cut
  using bars
")
```

### Stacked bars

```{r bar-stacked}
dbGetPlot(con, "
  visualize
    cut as x,
    count(*) as y,
    color as color
  from diamonds
  group by
    cut, color
  using bars
")
```

### Unstacked (dodged) bars

```{r bar-unstacked}
dbGetPlot(con, "
  visualize
    cut as x,
    count(*) as y,
    color as color
  from diamonds
  group by
    cut, color
  using unstacked bars
")
```

## Histograms

### Basic histogram

```{r hist-basic}
dbGetPlot(con, "
  visualize
    bin(mpg) as x,
    count(*) as y
  from cars
  group by
    bin(mpg)
  using bars
")
```

### Stacked histogram by group

```{r hist-grouped}
dbGetPlot(con, "
  visualize
    bin(mpg) as x,
    count(*) as y,
    cyl_cat as color
  from (
    select *, cast(cyl as varchar) as cyl_cat
    from cars
  )
  group by
    bin(mpg), cyl_cat
  using bars
")
```

## Line plots

### Single line

```{r line-single}
dbGetPlot(con, "
  visualize
    age as x,
    circumference as y
  from trees
  using line
")
```

### Multiple lines with collect by

```{r line-multiple}
dbGetPlot(con, "
  visualize
    age as x,
    circumference as y
  from trees
  collect by
    Tree
  using lines
")
```

### Regression lines by group

```{r line-regression-grouped}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y,
    cyl as color
  from (
    select hp, mpg, cyl::varchar as cyl
    from cars
  )
  using regression lines
")
```

## Box plots

### Basic box plot

```{r box-basic}
dbGetPlot(con, "
  visualize
    cut as x,
    price as y
  from diamonds
  using boxes
")
```

## Pie charts

A pie chart is a stacked bar chart in polar coordinates. Map a count to `theta`
and a category to `color`:

```{r pie}
dbGetPlot(con, "
  visualize
    count(*) as theta,
    cut as color
  from diamonds
  group by
    cut
  using bars
")
```

## Faceted plots

### Horizontal facets

```{r facet-horizontal}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from cars
  using points
  facet by
    cyl
")
```

### Vertical facets

```{r facet-vertical}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from cars
  using points
  facet by
    cyl vertically
")
```

### Facet grid

```{r facet-grid}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from cars
  using points
  facet by
    cyl,
    am
")
```

## Log-scaled plots

```{r log-scale}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from cars
  using (
    points
    layer
    regression line
  )
  scale by
    log(x),
    log(y)
")
```

## Multi-layer plots

### Points and regression line

```{r multi-layer-basic}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from cars
  using points

  layer

  visualize
    hp as x,
    mpg as y
  from cars
  using regression line
")
```

### Custom titles

```{r multi-with-titles}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y,
    cyl as color
  from cars
  using points
  title
    x as 'Horsepower',
    y as 'Miles Per Gallon',
    color as 'Cylinders'
")
```

```{r cleanup, include = FALSE}
DBI::dbDisconnect(con, shutdown = TRUE)
```
