---
title: "Get Started with rsgl"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with rsgl}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## What is SGL?

SGL (Structured Graphics Language) is a declarative language for generating
statistical graphics from relational data. It is designed to feel like SQL — if
you can write a `SELECT` statement, you can write a SGL statement.

rsgl implements SGL as an R package. You pass a SGL statement and a DuckDB
connection to `dbGetPlot()`, and it returns a ggplot2 plot object.

## Setup

Install rsgl from GitHub:

```r
# install.packages("remotes")
remotes::install_github("sgl-projects/rsgl")
```

Load the package and create a DuckDB connection with some data:

```{r setup}
library(rsgl)
library(duckdb)

con <- dbConnect(duckdb())
dbWriteTable(con, "cars", mtcars)
```

The `cars` table now contains the `mtcars` dataset with columns like `hp`
(horsepower), `mpg` (miles per gallon), `cyl` (cylinders), and `wt` (weight).

## Your first plot

A SGL statement has three required parts:

- **`visualize`** — maps columns to visual properties (aesthetics)
- **`from`** — specifies the data source
- **`using`** — chooses the geometric object (geom)

```{r first-plot}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from cars
  using points
")
```

This maps `hp` to the x-axis and `mpg` to the y-axis, pulls data from the
`cars` table, and draws a point for each row.

## Adding color

Map a third column to the `color` aesthetic to distinguish groups:

```{r color}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y,
    cyl as color
  from cars
  using points
")
```

## Changing geoms

Swap `points` for a different geom to change the representation. Use `bars` for
a bar chart — here combined with `count(*)` to count rows per group:

```{r bars}
dbGetPlot(con, "
  visualize
    cyl as x,
    count(*) as y
  from cars
  group by
    cyl
  using bars
")
```

Use `line` for a line chart, and `boxes` for box plots:

```{r boxes}
dbGetPlot(con, "
  visualize
    cyl_cat as x,
    mpg as y
  from (
    select mpg, cast(cyl as varchar) as cyl_cat
    from cars
  )
  using boxes
")
```

## Layering

Combine multiple geoms with the `layer` keyword. This overlays a regression
line on a scatterplot:

```{r layers}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from cars
  using (
    points
    layer
    regression line
  )
")
```

## Next steps

The [SGL Language Guide](sgl-language-guide.html) covers the full syntax
including transformations, grouping, collection, scaling,
faceting, coordinate systems, and titles.

```{r cleanup, include = FALSE}
DBI::dbDisconnect(con, shutdown = TRUE)
```
