---
title: "Get started with rsgl"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with rsgl}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This tutorial introduces the SGL language
as well as usage of the rsgl package.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup

For use with the examples in this tutorial, we will create an in-memory
[DuckDB](https://duckdb.org) database and load it with two tables,
`cars` and `trees`.
```{r setup}
library(rsgl)
library(duckdb)

con <- dbConnect(duckdb())
dbWriteTable(con, "cars", cars)
dbWriteTable(con, "trees", trees)
```

Let's query each to view a sample of data:

```{r cars_sample_query}
dbGetQuery(con, "
  select *
  from cars
  limit 5
")
```

```{r trees_sample_query}
dbGetQuery(con, "
  select *
  from trees
  limit 5
")
```

## dbGetPlot

The primary interface to rsgl is the `dbGetPlot` function,
which takes a [DBI](https://dbi.r-dbi.org) database connection
and a SGL statement and returns a [ggplot2](https://ggplot2.tidyverse.org) plot.
Although the examples in this tutorial use a connection to
a DuckDB database, `dbGetPlot` will accept any DBI connection.

## The SGL Language

### The From Clause

The `from` keyword precedes a data source specification, which is often the name of a table in the database.
Here, we specify the `cars` table as the data source.

```{r from_cars}
dbGetPlot(con, "
  visualize
    horsepower as x,
    miles_per_gallon as y
  from cars
  using points
")
```

This is similar in usage to the `from` keyword in SQL,
except that only a single data source is allowed (i.e., a comma-separated list of table names is not valid). 
If data from multiple sources or pre-processing of data is necessary,
then a SQL subquery can be provided:

```{r, from_subquery}
dbGetPlot(con, "
  visualize
    horsepower as x,
    miles_per_gallon as y
  from (
    select *
    from cars
    where origin = 'Japan'
  )
  using points
")
```

### The Using Clause

The `using` keyword precedes the name of the geometric object(s) that will represent the data.
Following ggplot2 terminology, these geometric objects are referred to as geoms.
Our previous examples demonstrated representing data with point geoms.

### The Visualize Clause

The `visualize` keyword precedes the aesthetic-to-column mapping,
which maps perceivable traits of the geoms to data source columns.
Our prior examples mapped the `x` and `y` positions of the point geoms to data source columns.
However, aesthetics may be non-positional, as shown below.
The `visualize` keyword most closely resembles the `select` keyword within SQL.

```{r, visualize}
dbGetPlot(con, "
  visualize
    horsepower as x,
    miles_per_gallon as y,
    origin as color
  from cars
  using points
")
```

In addition to mapping aesthetics to columns,
aesthetics can be mapped to expressions that include transformations and aggregations.

### Column-Level Transformations and Aggregations

SGL supports column-level transformations and aggregations,
as shown below where a binning transformation is combined with a count aggregation
to produce a histogram on `miles_per_gallon`.

```{r, histogram}
dbGetPlot(con, "
  visualize
    bin(miles_per_gallon) as x,
    count(*) as y
  from cars
  group by
    bin(miles_per_gallon)
  using bars
")
```

Here we see that, similarly to SQL,
SGL has a `group by` clause where aggregation groupings are specified.

Although SQL itself supports column-level transformation, grouping, and aggregation,
it is desirable for SGL to provide additional support for these operations.
Statistical graphics often require operations such as binning that are not natively supported by SQL.
Additionally, SGL's column-level transformations and aggregations are performed after scaling,
which cannot easily be replicated using SQL. Below is an example of this feature,
where binning and counting are applied after log scaling,
resulting in a log-scaled histogram.

```{r}
dbGetPlot(con, "
  visualize
    bin(miles_per_gallon) as x,
    count(*) as y
  from cars
  group by
    bin(miles_per_gallon)
  using bars
  scale by
    log(x)
")
```

### The Collect By Clause

In SGL, a geom is classified as individual if it represents each record (after transformation and aggregation) by a distinct geometric object.
Alternatively, a geom is classified as collective if it represents multiple records by one geometric object.
For example, points and lines are individual and collective geoms, respectively,
as shown below where the same data is represented using each.

```{r}
dbGetPlot(con, "
  visualize
    year as x,
    avg(miles_per_gallon) as y
  from cars
  group by
    year
  using points
")
```

```{r}
dbGetPlot(con, "
  visualize
    year as x,
    avg(miles_per_gallon) as y
  from cars
  group by
    year
  using line
")
```


For collective geoms, the collection of records to represent by each object
is determined automatically using reasonable defaults.
This behavior can be overridden by providing explicit collections in the `collect by` clause.
The `collect by` clause is similar to the `group by` clause,
except that rather than defining groups to aggregate by,
the `collect by` clause defines collections of records to be represented by one object.
Below we see an example where the default collection is not ideal,
followed by a preferrable explicit collection.

```{r}
dbGetPlot(con, "
  visualize
    age as x,
    circumference as y
  from trees
  using line
")
```

```{r}
dbGetPlot(con, "
  visualize
    age as x,
    circumference as y
  from trees
  collect by
    tree_id
  using lines
")
```
