---
title: "SGL language guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SGL language guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rsgl)
```

A reference for the SGL grammar — designed to look and feel like SQL — as
implemented by rsgl.

## Statement shape

*Placeholder.* The overall form: `visualize <aesthetics> from <source> using
<geom> [<graphic clauses>]`.

## Aesthetics

*Placeholder.* Position aesthetics (`x`, `y`), non-position aesthetics
(`color`, `fill`, `size`, ...), and how column references map to them.

## Sources

*Placeholder.* Tables and subqueries.

## Geoms

*Placeholder.* `points`, `bars`, `lines`, `boxes` (and the `box` alias),
and any representation modifiers.

## Column transformations and aggregations (CTAs)

*Placeholder.* `bin`, `count`, `avg`, and the implicit `identity` CTA.

## Graphic clauses

*Placeholder.* `scale`, `facet`, and `title` — order-independent after
`using`.

## Errors

*Placeholder.* What parser and validation errors look like and how to read
them.
