---
title: "SGL Language Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SGL Language Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

This guide covers the full syntax of SGL (Structured Graphics Language) as
implemented in rsgl. For the formal language specification, see
[Chapman (2025)](https://doi.org/10.48550/arXiv.2505.14690).

## Setup

All examples use `dbGetPlot()`, which takes a DuckDB connection and an SGL
statement. We'll load some datasets into an in-memory DuckDB database to work
with.

```{r setup}
library(rsgl)
library(duckdb)

con <- dbConnect(duckdb())
dbWriteTable(con, "cars", mtcars)
dbWriteTable(con, "trees", as.data.frame(Orange))

diamonds <- ggplot2::diamonds
diamonds$cut <- as.character(diamonds$cut)
diamonds$color <- as.character(diamonds$color)
diamonds$clarity <- as.character(diamonds$clarity)
dbWriteTable(con, "diamonds", diamonds)
```

## Statement structure

An SGL statement is built from clauses. A minimal statement has three: `visualize`,
`from`, and `using`. Additional clauses control grouping, scaling, faceting, and
titles. The statement ends with a semicolon (optional when passed as a string to
`dbGetPlot()`).

```
visualize
  <aesthetic mappings>
from <data source>
[group by <grouping expressions>]
[collect by <collection expressions>]
using <geom expression>
[scale by <scale expressions>]
[facet by <facet expressions>]
[title <title expressions>]
```

## The `from` clause

The `from` clause specifies the data source. This is typically a table name:

```{r from-table}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from cars
  using points
")
```

A SQL subquery can be used for filtering or transforming data before
visualization. The subquery must be enclosed in parentheses:

```{r from-subquery}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from (
    select hp, mpg
    from cars
    where hp < 200
  )
  using points
")
```

## The `visualize` clause

The `visualize` clause maps data source columns to aesthetics. Each mapping has
the form `<column> as <aesthetic>`.

### Positional aesthetics

Cartesian coordinates use `x` and `y`:

```{r viz-cartesian}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from cars
  using points
")
```

Polar coordinates use `theta` (angle) and `r` (radius) — see the
[coordinate systems](#coordinate-systems) section for examples.

A single positional aesthetic is also valid:

```{r viz-one-dim}
dbGetPlot(con, "
  visualize
    mpg as x
  from cars
  using points
")
```

### Non-positional aesthetics

`color` maps a column to color:

```{r viz-color}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y,
    cyl as color
  from cars
  using points
")
```

`size` maps a column to point size:

```{r viz-size}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y,
    cyl as size
  from cars
  using points
")
```

Multiple aesthetics can be combined:

```{r viz-combined}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y,
    cyl as color,
    wt as size
  from cars
  using points
")
```

### Column expressions

Aesthetics can be mapped to expressions that include transformations or
aggregations, not just bare column names. The supported column-level
transformations and aggregations are covered in their own section below.

## The `using` clause

The `using` clause specifies which geometric object (geom) represents the data.
Both singular and plural forms are accepted as keywords.

### Points

`points` (or `point`) draws individual points — one per row:

```{r geom-points}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from cars
  using points
")
```

### Bars

`bars` (or `bar`) draws bar segments. Bars are stacked by default when a
non-positional grouping is present:

```{r geom-bars}
dbGetPlot(con, "
  visualize
    cut as x,
    count(*) as y,
    color as color
  from diamonds
  group by
    cut, color
  using bars
")
```

### Lines

`lines` (or `line`) draws connected lines. Lines are collective geoms — they
represent multiple rows with a single geometric object:

```{r geom-lines}
dbGetPlot(con, "
  visualize
    age as x,
    circumference as y
  from trees
  collect by
    Tree
  using lines
")
```

### Boxes

`boxes` (or `box`) draws box plots. Like lines, boxes are collective geoms:

```{r geom-boxes}
dbGetPlot(con, "
  visualize
    cut as x,
    price as y
  from diamonds
  using boxes
")
```

## Column-level transformations and aggregations

SGL supports transformations and aggregations as part of column expressions in
the `visualize` clause.

### `bin()`

The `bin()` transformation groups a continuous column into discrete bins:

```{r cta-bin}
dbGetPlot(con, "
  visualize
    bin(mpg) as x,
    count(*) as y
  from cars
  group by
    bin(mpg)
  using bars
")
```

### `count()`

The `count(*)` aggregation counts rows per group. Any non-aggregated column in
the `visualize` clause must also appear in the `group by` clause — the same
rule as SQL:

```{r cta-count}
dbGetPlot(con, "
  visualize
    cut as x,
    count(*) as y
  from diamonds
  group by
    cut
  using bars
")
```

### Combining transformations and aggregations

`bin()` and `count(*)` can be combined to produce histograms. Non-positional
groupings create stacked histograms:

```{r cta-combined}
dbGetPlot(con, "
  visualize
    bin(mpg) as x,
    count(*) as y,
    cyl_cat as color
  from (
    select
      *,
      cast(cyl as varchar) as cyl_cat
    from cars
  )
  group by
    bin(mpg),
    cyl_cat
  using bars
")
```

## The `group by` clause

The `group by` clause specifies grouping expressions for aggregations. It
follows the same semantics as SQL's `GROUP BY`: any non-aggregated expression in
the `visualize` clause must appear in the `group by` clause.

```{r group-by}
dbGetPlot(con, "
  visualize
    cut as x,
    count(*) as y
  from diamonds
  group by
    cut
  using bars
")
```

## The `collect by` clause

Collective geoms (lines and boxes) represent multiple rows with a single
geometric object. By default, the collection of rows is determined implicitly.
The `collect by` clause overrides this behavior, explicitly specifying which
column determines how rows are grouped into separate geometric objects.

Without `collect by`, all rows feed into a single line:

```{r collect-default}
dbGetPlot(con, "
  visualize
    age as x,
    circumference as y
  from trees
  using line
")
```

With `collect by`, each unique value of the specified column produces a separate
line:

```{r collect-explicit}
dbGetPlot(con, "
  visualize
    age as x,
    circumference as y
  from trees
  collect by
    Tree
  using lines
")
```

## Geom qualifiers

Geom qualifiers are keywords placed before the geom name in the `using` clause.
They modify how the geom represents data.

### `jittered` (points)

Adds random positional variation to points, useful when data has overlapping
values:

```{r qual-jittered}
set.seed(42)
dbGetPlot(con, "
  visualize
    cyl_cat as x,
    mpg as y
  from (
    select mpg, cast(cyl as varchar) as cyl_cat
    from cars
  )
  using jittered points
")
```

### `regression` (lines)

Fits a linear regression line instead of connecting data points:

```{r qual-regression}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from cars
  using regression line
")
```

### `unstacked` (bars)

Positions bars side-by-side instead of stacking them:

```{r qual-unstacked}
dbGetPlot(con, "
  visualize
    cut as x,
    count(*) as y,
    color as color
  from diamonds
  group by
    cut, color
  using unstacked bars
")
```

## The `layer` operator

Multiple layers of geometric objects can be combined into a single graphic using
the `layer` keyword. Each layer is a complete sub-statement with its own
`visualize`, `from`, and `using` clauses:

```{r layer-full}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from cars
  using points

  layer

  visualize
    hp as x,
    mpg as y
  from cars
  using regression line
")
```

### Layered geom expressions

When layers share the same data source and aesthetic mapping, a shorthand syntax
avoids repetition. Geom expressions are listed inside parentheses in the `using`
clause, separated by `layer`:

```{r layer-shorthand}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from cars
  using (
    points
    layer
    regression line
  )
")
```

## The `scale by` clause

Each mapped aesthetic has a scale that determines how data values are mapped to
visual properties. Scales default to linear but can be overridden with the
`scale by` clause.

### Log scale

`log()` applies a base-10 logarithmic scale to an aesthetic:

```{r scale-log}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from cars
  using (
    points
    layer
    regression line
  )
  scale by
    log(x),
    log(y)
")
```

Note that scaling is applied to the aesthetic, not to the data. This means the
regression in the example above is computed on the log-scaled values. To
transform the data itself, use SQL in the `from` clause:

```{r scale-vs-transform}
dbGetPlot(con, "
  visualize
    log_hp as x,
    log_mpg as y
  from (
    select
      log(hp) as log_hp,
      log(mpg) as log_mpg
    from cars
  )
  using (
    points
    layer
    regression line
  )
")
```

## Coordinate systems

The coordinate system is inferred from the positional aesthetics in the
`visualize` clause. `x` and `y` imply Cartesian coordinates. `theta` and `r`
imply polar coordinates.

A pie chart is a stacked bar chart rendered in polar coordinates:

```{r coord-pie}
dbGetPlot(con, "
  visualize
    count(*) as theta,
    cut as color
  from diamonds
  group by
    cut
  using bars
")
```

## The `facet by` clause

Faceting generates small multiples — separate panels for each unique value of a
column. A single facet expression produces horizontal panels by default:

```{r facet-single}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from (
    select
      *,
      cast(cyl as varchar) as cyl_cat
    from cars
  )
  using points
  facet by
    cyl_cat
")
```

The orientation can be changed with the `vertically` keyword:

```{r facet-vertical}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from (
    select
      *,
      cast(cyl as varchar) as cyl_cat
    from cars
  )
  using points
  facet by
    cyl_cat vertically
")
```

Two facet expressions produce a grid — one varies horizontally, the other
vertically:

```{r facet-grid}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y
  from (
    select
      *,
      cast(cyl as varchar) as cyl_cat,
      cast(am as varchar) as am_cat
    from cars
  )
  using points
  facet by
    cyl_cat,
    am_cat
")
```

## The `title` clause

Axis and legend titles are automatically derived from aesthetic mappings. The
`title` clause overrides these defaults:

```{r title}
dbGetPlot(con, "
  visualize
    hp as x,
    mpg as y,
    cyl as color
  from cars
  using points
  title
    x as 'Horsepower',
    y as 'Miles Per Gallon',
    color as 'Cylinders'
")
```

```{r cleanup, include = FALSE}
DBI::dbDisconnect(con, shutdown = TRUE)
```
